Chaotic mixing of tidal debris

### 2015-06-17 ms.tex
words yo

% packages
\usepackage{amssymb,amsmath,amsbsy}
\usepackage{booktabs}
\usepackage{bbold}
\usepackage{mathrsfs}
\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage[yyyymmdd]{datetime}
\renewcommand{\dateseparator}{-}
\usepackage{lscape}
\usepackage{rotating}

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or ``Perfect Ellipsoid'' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change the sense of rotation about any axis.

The frequencies of a generic, non-resonant orbit are incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} These conditionally-periodic orbits uniformly cover the surface of an orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ the orbit is a resonant orbit. For resonant orbits, the existence of an additional \emph{local} constant of motion reduces the dimensionality of the orbit. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. For a triaxial potential the resonances define planes in frequency-space, but resonance surfaces in action-space are generic 2D surfaces. In plots of frequency ratios, the resonances appear as lines; Figure~\ref{fig:cartoons} (left panel) shows a cartoon portrait of a portion of frequency-space for an integrable potential with resonance lines, non-resonant, and resonant orbits marked. For an integrable potential, all non-resonant orbits are regular.

@@

The orbit structure of near-integrable potentials can be understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $H_0(\bs{J})$ represents an integrable Hamiltonian, and $\epsilon$ is a small parameter that determines the perturbation strength \citep[a complete description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become ``thick'' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99}, and these layers regions are then generically surrounded by stochastic layers where chaotic motion occurs. Chaotic orbits are characterized by exhibiting random behavior despite being generated by entirely deterministic equations of motion. Stochastic layers form gaps in action space where tori are destroyed; the orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies are not constant and diffuse chaotically with time.

Figure~\ref{fig:cartoons} (middle panel) shows a cartoon of frequency space for a near-integrable potential (a small perturbation away from the potential in the left panel). Much of the structure that was present in the integrable potential remains in the near-integrable case, but the differences are highlighted. Orbits in the resonant layers surrounding the resonances (grey) are near-resonant orbits that will librate around the resonance. Chaotic orbits in the stochastic layer (red) behave erratically depending on the surrounding resonance structure. If the stochastic layer is small and the chaotic orbit is therefore confined in some way, it may behave nearly regular for long periods of time.

@@

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori caused by the onset of chaotic motion.

@@

Chaotic orbits ...

@@

Our goal is to map the orbit structure of arbitrary potentials, with an emphasis on identifying the chaotic orbits and and understanding the evolution of these orbits over short times. In this section, we describe the methods we will use to detect and quantify the strength of chaos for large grids of orbits.

@@

 \citet[][hereafter JS02]{jing02} found that a triaxial generalization of the NFW density profile \citep{navarro96} generates excellent fits to the density distributions within haloes in their high-resolution N-body simulations, and they provide probability distributions for the axis ratios of a large sample of these halos. JS02 finds median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that JS02 use the opposite notation so that $c$ is the major and $a$ is the minor axis.} These are largely consistent with findings from more recent simulations \citep[e.g.,][]{millenium, hmm, butsky15} and consistent with constraints from weak lensing that place a lower limit on minor-to-major axis ratios of $c/a\gtrsim0.5$ \citep{vanuitert12}. JS02 finds significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

The MLE, $\lambda_{\rm max}$, is interpreted as a rate that sets the exponential divergence of infinitesimally close chaotic orbits. It is therefore useful to consider the corresponding $e$-folding time, 
\begin{equation}
	t_{\rm \lambda} = \frac{1}{\lambda_{\rm max}}.
\end{equation}
We will use this as the prediction from the Lyapunov exponent for the timescale over which chaos should be dynamically important for a given orbit.

@@

Bounded, regular orbits in a triaxial potential have three fundamental frequencies, $\bs{\Omega}$, that determine the periodic behavior of motion. The motion in any canonical coordinate can therefore be decomposed as a Fourier sum (Equation~\ref{eq:fourier}) where the Fourier frequencies are linear, integer combinations of the fundamental frequencies (Equation~\ref{eq:fourierfreq}). \cite{laskar93} introduced a method for recovering the fundamental frequencies of an orbit that effectively uses a Hamming-filtered fast-Fourier transform (FFT) of complex combinations of the motion (e.g., $x(t) + i v_x(t)$). It has been shown that using a filtered FFT makes the accuracy in determining the frequencies converge much faster than the typical $\inttime^{-1}$ expected for a standard FFT \citep{laskar99}, where $\inttime$ is the integration time. The FFT is iteratively searched for the strongest frequency component, which is subtracted from the spectrum after each iteration. This procedure generates a table of frequencies for each component of motion which must then be searched for the three fundamental frequencies.

@@

If an orbit is chaotic, the motion can no longer be expanded in terms of a single set of fundamental frequencies. For a weakly chaotic orbit, the orbit may appear consistently periodic over long windows of time. \superfreq\ will pick out a set of frequencies for such orbits that correspond to the largest peaks in the power spectrum of the stochastic orbits, however these peaks will change location and amplitude with time. For more strongly chaotic orbits, the power spectrum will be quite noisy and the peak frequencies may change erratically when comparing two separate sections of orbit. The frequencies picked out by \superfreq\ for such orbits will therefore represent the average periodic nature of the orbit over a given period. We define the fractional frequency diffusion rate per orbital period, $\mathcal{R}$, in the $i$th fundamental frequency as
\begin{equation}
	\mathcal{R}_i = \frac{\Omega_i^{(2)} - \Omega_{i}^{(1)}}{\Omega_{i}^{(1)} \, N} \label{eq:fdrate}
\end{equation}
where $\Omega_i^{(1)}$ and $\Omega_i^{(2)}$ is the $i$th fundamental frequency computed in each of two consecutive sections of orbit and $N$ is the number of orbital periods in a single section of orbit. By inverting this rate, we can compute the timescale over which we expect order-unity changes to the fundamental frequencies: the \emph{frequency diffusion time} is defined as
\begin{equation}
	t_\Omega = \, (\max_{a_i} \, \mathcal{R}_i)^{-1} \label{eq:fdtime}
\end{equation}
where the maximum is taken with respect to the corresponding amplitudes, $a_i$, of the fundamental frequency components \citep[see][]{valluri12}, and the time is in units of number of orbital periods, ${\rm T}$. For a small ensemble of orbits (e.g., tidal debris), the relevant timescale is the time over which the change in frequencies for a single orbit is comparable to the spread of frequencies in the ensemble. We can estimate this timescale by multiplying the frequency diffusion time by a factor equal to the fractional spread in frequencies of the debris. For example, a globular cluster typically has $\approx$1\% spreads in fundamental frequencies, so by multiplying the frequency diffusion time by $f = 0.01$, we can estimate the time (number of orbital periods) over which we expect the frequencies to evolve by this amount.

@@

We generate isoenergy grids of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass) chosen to span a range of distances comparable to the scale radius of the potential ($E=-(397.2~{\rm km~s}^{-1})^2$ in physical units; see Table~\ref{tbl:potential}). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits --- the other major class of orbits in triaxial potentials --- which tend to plunge deep into the inner regions of the potential, however there are appreciable regions of stochastic tube orbits. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the faster disruption expected for stellar systems on radially plunging orbits. The major classes of stable tube orbits circulate about either the major or minor axis. This grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits), but the most numerous orbits in this grid are the short-axis and outer long-axis tubes.

To compute the Lyapunov exponents, we integrate all orbits in the grid for 1000 orbital periods \todo{apw}{re-run for 5000 or 10000 periods?} and use the method described in Section~\ref{sec:lyap}. Figure~\ref{fig:lyapmap} shows the grid of initial conditions (in the $xz$ plane) --- each pixel corresponds to an orbit, and the logarithm of the Lyapunov time (in units of Gyr \todo{apw}{ convert to num. periods?}) is mapped to greyscale intensity. The darker pixels have shorter Lyapunov times and are more chaotic. By fixing the integration time (in units of orbital periods), the MLE cannot detect weak chaos and the majority of orbits appear to be regular because they have exceedingly long Lyapunov times (all white points have $t_\lambda \gtrsim XX$).

For each orbit, we also separately integrate for $\approx$80 orbital periods and use \superfreq\ to compute the fundamental frequencies for the two consecutive sections of 40 orbital periods. We have found that this is the shortest integration window for which we are able to consistently and accurately recover the frequencies using \superfreq. Figure~\ref{fig:freqdiff} shows the same grid of initial conditions, but now the greyscale intensity is set by the logarithm of the (order-unity) frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of \superfreq\ --- weak chaos (grey) is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit of interest, or of how long it takes for the frequencies to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. Does the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent?

@@

It is computationally intractable to run full N-body simulations for the large grid of orbital initial conditions of the previous section and we therefore take a simplified approach for studying how finite-volume debris spreads along each of these orbits. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions --- the ``parent'' orbit --- we find the phase-space position of the nearest pericenter by integrating forward for a short time until a minimum is found, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all test particles for a specified integration time. The physical scale of the ensemble is set by the tidal radius in position and the velocity scale in velocity and are therefore set by the mass of the progenitor \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M_{\rm enc}(\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M_{\rm enc}(r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. We take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass of the host potential to estimate the above debris scales. Figure~\ref{fig:ensembles} shows such ensembles initialized around the three orbits described above and evolved for 16 orbital periods (same as the N-body simulations); the morphology of the ensembles is visually similar to the ``oldest'' (first stripped) debris in the N-body simulations.

@@

Why does chaos manifest itself after just 16 orbital periods around orbits with predicted chaotic timescales equal to hundreds or thousands of orbital periods? To understand this, we first return to the concepts of nonlinear dynamics. In near-integrable potentials, chaotic diffusion of an orbit in actions and frequencies occurs both across resonance layers (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion). For weakly chaotic orbits, motion across a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is comparable (within a factor of a few) to the orbital frequency, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can't change by large factors (unless there is resonance overlap and the motion is strongly chaotic). The rate of diffusion along stochastic layers via Arnold diffusion depends on the local resonant structure and is hard to predict, however, this has been done analytically for simple potentials \citep[e.g.,][]{chirikov??}. For systems with $N>2$ dof, the stochastic layers connect and form an intricate network of stochasticity known as the Arnold web; an orbit that ergodically mixes over its energy hypersurface must traverse this web, though the timescales typical for this phenomenon are many thousands of orbital periods.

>>>>

% packages
\usepackage{amssymb,amsmath,amsbsy}
\usepackage{booktabs}
\usepackage{bbold}
\usepackage{mathrsfs}
\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage[yyyymmdd]{datetime}
\renewcommand{\dateseparator}{-}
%\usepackage{lscape}
%\usepackage{rotating}

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or ``Perfect Ellipsoid'' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change sense of rotation about any axis.

The frequencies of a generic, non-resonant orbit are incommensurable --- that is, $\bs{n} \cdot \bs{\Omega} \neq 0$ for any integer vector, $\bs{n}$, with reasonable magnitude.\footnote{A more precise condition is stated in terms of a diophantine condition, e.g., $|\bs{n} \cdot \boldsymbol{\Omega}| > \alpha \, |n|^{-\gamma}$ where $\alpha, \gamma>0$.} These conditionally-periodic orbits uniformly cover the surface of an orbital torus. If instead there exists a relation of the form $\boldsymbol{n} \cdot \boldsymbol{\Omega} = 0$ the orbit is a resonant orbit. For resonant orbits, the existence of an additional \emph{local} constant of motion reduces the dimensionality of the orbit. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits, and if an additional resonance relation exists, \emph{bi-resonant}. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. For a triaxial potential the resonances define planes in frequency-space, but resonance surfaces in action-space are generic surfaces. In plots of frequency ratios, the resonances appear as lines; Figure~\ref{fig:cartoons} (left panel) shows a cartoon portrait of a portion of frequency-space for an integrable potential with example resonance lines, non-resonant, and resonant orbits marked. For an integrable potential, all non-resonant orbits are regular.

@@

The orbit structure of near-integrable potentials can be understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $H_0(\bs{J})$ represents an integrable Hamiltonian, and $\epsilon$ is a small parameter that determines the perturbation strength \citep[a description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become ``thick'' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99}. These resonant layers are then generically surrounded by stochastic layers where chaotic motion occurs. Chaotic orbits are characterized by exhibiting random behavior despite being generated by entirely deterministic equations of motion. Stochastic layers form gaps in action space where tori are destroyed; the orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies are not constant and diffuse chaotically with time.

Figure~\ref{fig:cartoons} (middle panel) shows a cartoon of frequency space for a near-integrable potential (a small perturbation away from the potential in the left panel). Much of the structure that was present in the integrable potential remains in the near-integrable case, but the differences are highlighted. Orbits in the resonant layers surrounding the resonances (grey) are near-resonant orbits that will librate around the resonance. Chaotic orbits in the stochastic layer (red) behave erratically depending on the surrounding resonance structure. If the stochastic layer is small and the chaotic orbit is therefore confined, the orbit may behave nearly regular for long periods of time.

@@

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori due to chaotic motion.

@@

Chaotic orbits have no orbital actions because they are only strictly bound to their energy hypersurface (if the potential is time-independent). The orbits therefore do not have a single set of fundamental frequencies, but rather the frequencies that describe the character of motion evolve with time. In near-integrable potentials, chaotic diffusion of an orbit in actions and frequencies occurs both across resonance layers (a sort of stochastic libration) and along the stochastic layers that surround the resonances (Arnold diffusion). For weakly chaotic orbits, motion across a resonance layer can occur with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is comparable (within a factor of a few) to the orbital frequency, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can't change by large factors (unless there is resonance overlap and the motion is strongly chaotic).

The rate of diffusion along stochastic layers via Arnold diffusion depends on the local resonant structure and is hard to predict, however, this has been done analytically for simple potentials \citep[e.g.,][]{chirikov??}. For systems with $N>2$ dof, the stochastic layers connect and form an intricate network of stochasticity known as the Arnold web; an orbit that ergodically mixes over its energy hypersurface must traverse this web, though the timescales typical for this phenomenon are many thousands of orbital periods. For galactic timescales, Arnold diffusion is not expected to be relevant, however chaotic motion across resonances can occur over short times. [the volume allowed to explore over short times is small, but could be relevant for tidal debris...]

\subsection{Mixing of orbit ensembles}

@@

Our goal is to map the orbit structure of arbitrary potentials, with an emphasis on identifying the chaotic orbits and understanding the evolution of these orbits over short times. In this section, we describe the methods we will use to detect and quantify the strength of chaos for large grids of orbits.

@@

 \citet[][hereafter JS02]{jing02} found that a triaxial generalization of the NFW density profile \citep{navarro96} generates excellent fits to the density distributions within haloes in their high-resolution (dark-matter-only) N-body simulations, and they provide probability distributions for the axis ratios of a large sample of these halos. JS02 finds median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that JS02 use the opposite notation so that $c$ is the major and $a$ is the minor axis.} These are largely consistent with findings from more recent simulations \citep[e.g.,][]{millenium, hmm, butsky15} and consistent with constraints from weak lensing that place a lower limit on minor-to-major axis ratios of $c/a\gtrsim0.5$ \citep{vanuitert12}. JS02 finds significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

The MLE, $\lambda_{\rm max}$, is interpreted as a rate that sets the exponential divergence of infinitesimally close chaotic orbits. It is therefore useful to consider the corresponding $e$-folding time, 
\begin{equation}
	t_{\rm \lambda} = \frac{1}{\lambda_{\rm max}} \todo{apw}{divide by T?}.
\end{equation}
We will use this as the prediction from the Lyapunov exponent for the timescale over which chaos should be dynamically important for a given orbit.

@@

Bounded, regular orbits in a triaxial potential have three fundamental frequencies, $\bs{\Omega}$, that determine the periodic behavior of motion. The motion in any canonical coordinate can therefore be decomposed as a Fourier sum (Equation~\ref{eq:fourier}) where the Fourier frequencies are linear, integer combinations of the fundamental frequencies (Equation~\ref{eq:fourierfreq}). \cite{laskar93} introduced a method for recovering the fundamental frequencies of an orbit that effectively uses a filtered fast-Fourier transform (FFT) of complex combinations of the motion (e.g., $x(t) + i v_x(t)$). It has been shown that using a filtered FFT makes the accuracy in determining the frequencies converge much faster than the typical $\inttime^{-1}$ expected for a standard FFT \citep{laskar99}, where $\inttime$ is the integration time. The FFT is iteratively searched for the strongest frequency component, which is subtracted from the spectrum after each iteration. This procedure generates a table of frequencies for each component of motion which must then be searched for the three fundamental frequencies.

@@

If an orbit is chaotic, the motion can no longer be expanded in terms of a single set of fundamental frequencies. For a weakly chaotic orbit, the orbit may appear consistently periodic over long windows of time. \superfreq\ will pick out a set of frequencies for such orbits that correspond to the largest peaks in the power spectrum of the stochastic orbits, however these peaks will change location and amplitude with time. For more strongly chaotic orbits, the power spectrum will be quite noisy and the peak frequencies may change erratically when comparing two separate sections of orbit. The frequencies picked out by \superfreq\ for such orbits will therefore represent the average periodic nature of the orbit over a given integration window. We define the fractional frequency diffusion rate per orbital period, $\mathcal{R}$, in the $k$th fundamental frequency as
\begin{equation}
	\mathcal{R}_k = \frac{\Omega_{k}^{(2)} - \Omega_{k}^{(1)}}{\Omega_{k}^{(1)} \, N} \label{eq:fdrate}
\end{equation}
where the upper index refers to which of the two consecutive sections of orbit and $N$ is the number of orbital periods in a single section of orbit. By inverting this rate, we can compute the timescale over which we expect order-unity changes to the fundamental frequencies: the \emph{frequency diffusion time} is defined as
\begin{equation}
	t_\Omega = \, (\max_{a_k} \, \mathcal{R}_k)^{-1} \label{eq:fdtime}
\end{equation}
where the maximum is taken with respect to the corresponding amplitudes, $a_i$, of the fundamental frequency components \citep[see][]{valluri12}, and the time is in units of number of orbital periods, ${\rm T}$.

For a small ensemble of orbits (e.g., tidal debris), the relevant timescale is the time over which the change in frequencies for a single orbit is comparable to the spread of frequencies in the ensemble. We can estimate this timescale by multiplying the frequency diffusion time by a factor equal to the fractional spread in frequencies of the debris. For example, a globular cluster typically has $\approx$1\% spreads in fundamental frequencies, so by multiplying the frequency diffusion time by $f = 0.01$, we can estimate the time (number of orbital periods) over which we expect the frequencies to evolve by this amount.

@@

We generate isoenergy grids of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass), $E$, chosen to span a range of distances comparable to the scale radius of the potential ($E=-(397.2~{\rm km~s}^{-1})^2$ in physical units; see Table~\ref{tbl:potential}). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits, which tend to plunge deep into the inner regions of the potential where the density profile is steep, however there are appreciable regions of stochastic tube orbits. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the faster disruption expected for stellar systems on radially plunging orbits. The major classes of stable tube orbits circulate about either the major or minor axis. This grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits), but the most numerous orbits in this grid are the short-axis and outer long-axis tubes.

We integrate all orbits in the grid for 1000 orbital periods \todo{apw}{re-run for 5000 or 10000 periods?} and use the method described in Section~\ref{sec:lyap} to compute the Lyapunov exponents. Figure~\ref{fig:lyapmap} shows the grid of initial conditions (in the $xz$ plane) --- each pixel corresponds to an orbit, and the logarithm of the Lyapunov time (in units of Gyr \todo{apw}{ convert to num. periods?}) is mapped to greyscale intensity. The darker pixels have shorter Lyapunov times and are more chaotic. By fixing the integration time (in units of orbital periods), the MLE cannot detect weak chaos and the majority of orbits appear to be regular because they have exceedingly long Lyapunov times (all white points have $t_\lambda \gtrsim XX$). \todo{apw}{describe other features in lyapunov grid?}

For each orbit, we also separately integrate for $\approx$80 orbital periods and use \superfreq\ to compute the fundamental frequencies for the two consecutive sections of 40 orbital periods. We have found that this is the shortest integration window for which we are able to consistently and accurately recover the frequencies using \superfreq. Figure~\ref{fig:freqdiff} shows the same grid of initial conditions, but now the greyscale intensity is set by the logarithm of the (order-unity) frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of \superfreq\ --- weak chaos (grey) is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs. \todo{apw}{describe unstable resonances, etc.}

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit of interest, or of how long it takes for the frequencies of a single orbit to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. Does the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent?

@@

It is computationally intractable to run full N-body simulations for the large grid of orbital initial conditions of the previous section and we therefore take a simplified approach for studying how finite-volume debris spreads along each of these orbits. We instead consider small ensembles of particles meant to represent debris disrupted from a single tidal disruption event. For a given set of orbital initial conditions --- the ``parent'' orbit --- we find the phase-space position of the nearest pericenter by integrating forward for a short time until a minimum is found, initialize a small ensemble of test particle orbits around this position, then follow the orbits of all test particles for a specified integration time. The physical scale of the ensemble is set by the tidal radius in position and the velocity scale in velocity and are therefore set by the mass of the progenitor \citep[e.g.,][]{johnston98, apw14}. If $(\bs{x}_0,\bs{v}_0)$ are the parent orbit initial conditions at pericenter, then $(\delta\bs{x}_i,\delta\bs{v}_i)$ is the deviation vector of the $i$th particle and the magnitude of the offsets are assumed to be Normally distributed away from the parent orbit:
\begin{align}
	\delta\bs{x}_i &\sim \mathcal{N}(0, r_{\rm tide})\\
	\delta\bs{v}_i &\sim \mathcal{N}(0, \sigma_v)\\
	r_{\rm tide} &= \|\bs{x}_0\| \left(\frac{m}{3M(<\|\bs{x}_0\|)}\right)^{1/3} \\
	\sigma_v &= \|\bs{v}_0\|\left(\frac{m}{3M(<\|\bs{x}_0\|)}\right)^{1/3} 
\end{align}
where $M(<r)$ is the mass enclosed of the host potential within radius $r$, $m$ is the mass scale of the debris, and $\|\cdot \|$ is the Euclidean norm. We take $m=10^4~\msun$ to represent globular-cluster-like progenitors, and use the spherically-averaged enclosed mass of the host potential to estimate the above debris scales. Figure~\ref{fig:ensembles} shows final positions of test-particle0 ensembles initialized around the three orbits described above and evolved for 16 orbital periods (same as the N-body simulations); the morphology of the ensembles is visually similar to the ``oldest'' (first stripped) debris in the N-body simulations.

@@

Why does chaos manifest itself after just 16 orbital periods around orbits with predicted chaotic timescales equal to hundreds or thousands of orbital periods? Both the Lyapunov exponent and the frequency diffusion rate measure mean, long-term rates of chaotic diffusion. If a weakly chaotic orbit is confined (by other nearby, non-overlapping resonances or stable regions), the mean drift of an orbit in frequency space may be small if computed over timescales long compared to the orbital time but short compared to the Arnold diffusion time. For the potential considered in this paper, it turns out that these are comparable to the timescales for which the frequency diffusion rate of the previous section were computed ($\approx$100-1000 orbital periods). This window length was chosen in order to have a strong enough signal in the frequency spectrum of chaotic orbits to measure the most significant frequencies using \superfreq\, however, it erases the small-scale variation of the frequencies.
